Retrofitting Effect Handlers onto OCaml

### 2020-11-01 retro-concurrency.tex
Updates

\begin{lstlisting}[numbers=left]
type sign = P | N
type val = Z of int | Error | Inf of sign
type exp = Val of val | ... | Div of exp*exp
effect DivBy0 : int -> val

let rec eval e =
  match e with
  | Val v -> v
  | ...
  | Div (e1,e2) ->
    match eval e1, eval e2 with
    | Z v1, Z v2 ->
        if v2 = 0 then perform (DivBy0 v1)
        else Z (v1/v2)
    | Error, _ -> Error
    | Inf s, _ -> Inf s
    | ...

let (/) n d = Div (Val (Z n), Val (Z d))
\end{lstlisting}

@@

\begin{lstlisting}
let ios e = match eval e with
| v -> v (* value case *)
| effect (DivBy0 _) k -> (* effect case *)
    continue k Error
\end{lstlisting}

@@

\begin{lstlisting}
type exp = ... | Inp
effect Read : in_channel -> string

let rec eval e = match e with
| ...
| Inp -> int_of_string(perform(Read stdin))
\end{lstlisting}

@@

\begin{lstlisting}
let run e = match gsearch e with
| v -> v
| effect (Read chan) k ->
  register_async_input_line (chan, k);
  let ui = get_ui_k () in
  continue ui ()
\end{lstlisting}

@@

OCaml also generates precise DWARF unwind information for OCaml, thanks to
which debuggers such as |gdb| and |lldb|, and profilers such as |perf| work
out-of-the-box. For example, for the program in Figures~\ref{code:meander_ml}
and~\ref{code:meander_c}, one could set a break point in |gdb| at
|caml_raise_exn| to get the following backtrace.
\begin{lstlisting}[language=c,basicstyle=\ttfamily\footnotesize]
% gdb ./meander.exe
(gdb) break caml_raise_exn
Breakpoint 1 at 0x3e5dc
(gdb) r
Starting program: meander.exe
Breakpoint 1, 0x925dc in caml_raise_exn ()
(gdb) bt
#0  0x925dc in caml_raise_exn ()
#1  0x6fd3e in camlMeander__c_to_ocaml_83 () at meander.ml:5
#2  0x925a4 in caml_call_ocaml ()
#3  0x8a84a in caml_callback_exn (...) at callback.c:145
#4  caml_callback (...) at callback.c:199
#5  0x76e0a in ocaml_to_c (unit=1) at meander_.c:5
#6  0x6fd77 in camlMeander__omain_88 () at meander.ml:10
#7  0x6fe92 in camlMeander__entry () at meander.ml:14
#8  0x6f719 in caml_program ()
#9  0x925a4 in caml_call_ocaml ()
#10 0x92e4c in caml_startup_common (...) at startup_nat.c:162
#11 0x92eab in caml_startup_exn (...) at startup_nat.c:167
#12 caml_startup (...) at startup_nat.c:172
#13 0x6f55c in main (...) at main.c:44
\end{lstlisting}
\noindent which corresponds to the stack layout in
Figure~\ref{fig:stock_stack}.

@@

The interaction of non-local control flow with systems programming is quite
subtle~\cite{TFP17}. Consider the following function that uses blocking I/O
functions from the OCaml standard library to copy data from the input channel
|ic| to the output channel |oc|:

\begin{lstlisting}
let rec copy ic oc =
  try
    let l = input_line ic in
    output_string oc (l ^ "\n");
    copy ic oc
  with _ -> (close_in ic; close_out oc)
\end{lstlisting}

|input_line| raises |End_of_file| exception on reaching the end of input.
Although not apparent, the code is written in a defensive style to handle other
exceptional cases such as the channels being closed externally. Both
|input_line| and |output_string| raise |Sys_error| exception if the channel is
closed. This function cleanly closes the channels on exceptional cases;
|close_*| functions do nothing if the channel is already closed.

One of our goals(\S\ref{sec:requirements}) is to make this code transparently
asynchronous. We can define effects for performing the I/O operations:
\begin{lstlisting}
effect In_line : in_channel -> string
effect Out_str : out_channel -> string -> unit
\end{lstlisting}
\noindent and wrap them up in functions with the same signature as the one from
the standard library.
\begin{lstlisting}
let input_line ic = perform (In_line ic)
let output_string oc = perform (Out_str oc)
\end{lstlisting}

@@

For handling the effects, our implementation extends OCaml's |match ... with|
syntax with effect patterns. The expression
\begin{lstlisting}
match e with
| None -> false | Some b -> b
| effect (E s) k1 -> e1 | effect (F f) k2 -> e2
\end{lstlisting}
\noindent is translated to the equivalent of
\begin{lstlisting}
match_with (fun () -> e)
{ retc = (function None -> false | Some b -> b);
  effc = (function
  | (E s) -> (fun k1 -> e1)
  | (F f) -> (fun k2 -> e2)
  | e -> (fun k -> match perform e with
          | v -> continue k v
          | exception e -> discontinue k e)); }
\end{lstlisting}
For the sake of exposition, we introduce a |('a,'b) handler| type. This handler
handles a |'a comp| that returns a |'a| value, and itself returns a |'b| value.
The handler has a return case of type |'a -> 'b|. The effect case |effc|
handles effects of type |'c eff| with |('c,'b) continuation| and returns a
value of type |'b|. The last pattern in |effc| forwards any unmatched effect to
the outer handler and returns the value and exceptions back to the original
performer.

@@

Finally, we have the saved exception pointer, which points to the top-most
exception frame, and the saved stack pointer, which points to the top of the
stack. Switching between fiber only involves saving the exception and the stack
pointer of the current stack and loading the same on the target stack. We
allocate the space for the fiber using |malloc| and |free| it when no longer
required. We do not use a \emph{stack cache} of recently freed stacks for
speeding up fiber allocation. In our experiments, we observed that directly
using |mimalloc| outperforms having the stack cache.

@@

When we first enter OCaml from C, a new fiber is allocated for the main OCaml
stack. Since callbacks may be frequent in OCaml programs that use finalisers,
we run the callbacks on the same fiber as the current one. For example, the
layout of the Multicore OCaml stack at |caml_raise_exn| in the |meander|
example from \S\ref{sec:stack} is shown in Figure~\ref{fig:mcstack}. The
functions |caml_call_c| and |caml_call_ocaml| switch the stacks, and hence are
show in both the system stack and the fiber. The details of the exception
frames have been elided in the figure. Thanks to the fiber representation,
external calls and callbacks remain fast.

>>>>

\begin{lstlisting}[numbers=left]
type sign = P | N
type val = Z of int | Error | Inf of sign
type exp = Val of val | ... | Div of exp*exp
effect DivBy0 : int -> val
let rec eval e =
  match e with
  | Val v -> v
  | ...
  | Div (e1,e2) ->
    match eval e1, eval e2 with
    | Z v1, Z v2 ->
        if v2 = 0 then perform (DivBy0 v1)
        else Z (v1/v2)
    | Error, _ -> Error | Inf s, _ -> Inf s
    | ...
let (/) n d = Div (Val (Z n), Val (Z d))
\end{lstlisting}

@@

\begin{lstlisting}
let ios e = match eval e with
| v -> v
| effect (DivBy0 _) k -> continue k Error
\end{lstlisting}

@@

\begin{lstlisting}
type exp = ... | Inp
effect Read : in_channel -> string
let rec eval e = match e with
| ...
| Inp -> int_of_string(perform(Read stdin))
\end{lstlisting}

@@

\begin{lstlisting}
let run e = match gsearch e with
| v -> v
| effect (Read chan) k ->
  register_async_input_line (chan, k);
	continue (get_ui_k ()) ()
\end{lstlisting}

@@

OCaml also generates precise DWARF unwind information for OCaml, thanks to
which debuggers such as |gdb| and |lldb|, and profilers such as |perf| work
out-of-the-box. For example, for the program in Figures~\ref{code:meander_ml}
and~\ref{code:meander_c}, one could set a break point in |gdb| at
|caml_raise_exn| to get the following backtrace.
\begin{lstlisting}[language=c,basicstyle=\ttfamily\footnotesize]
#0  0x925dc in caml_raise_exn ()
#1  0x6fd3e in camlMeander__c_to_ocaml_83 () at meander.ml:5
#2  0x925a4 in caml_call_ocaml ()
#3  0x8a84a in caml_callback_exn (...) at callback.c:145
#4  caml_callback (...) at callback.c:199
#5  0x76e0a in ocaml_to_c (unit=1) at meander_.c:5
#6  0x6fd77 in camlMeander__omain_88 () at meander.ml:10
#7  0x6fe92 in camlMeander__entry () at meander.ml:13
#8  0x6f719 in caml_program ()
#9  0x925a4 in caml_call_ocaml ()
#10 0x92e4c in caml_startup_common (...) at startup_nat.c:162
#11 0x92eab in caml_startup_exn (...) at startup_nat.c:167
#12 caml_startup (...) at startup_nat.c:172
#13 0x6f55c in main (...) at main.c:44
\end{lstlisting}
\noindent which corresponds to the stack layout in
Figure~\ref{fig:stock_stack}.

@@

The interaction of non-local control flow with systems programming is quite
subtle~\cite{TFP17}. Consider the following function that uses blocking I/O
functions from the OCaml standard library to copy data from the input channel
|ic| to the output channel |oc|:
\begin{lstlisting}
let rec copy ic oc =
  try output_string oc ((input_line ic) ^ "\n");
      copy ic oc
  with _ -> (close_in ic; close_out oc)
\end{lstlisting}
|input_line| raises |End_of_file| exception on reaching the end of input.
Although not apparent, the code is written in a defensive style to handle other
exceptional cases such as the channels being closed externally. Both
|input_line| and |output_string| raise |Sys_error| exception if the channel is
closed. This function cleanly closes the channels on exceptional cases;
|close_*| functions do nothing if the channel is already closed.

One of our goals(\S\ref{sec:requirements}) is to make this code transparently
asynchronous. We can define effects for performing the I/O operations  and wrap
them up in functions with the same signature as the one from the standard
library:
\begin{lstlisting}
effect In_line : in_channel -> string
effect Out_str : out_channel -> string -> unit
let input_line ic = perform (In_line ic)
let output_string oc = perform (Out_str oc)
\end{lstlisting}

@@

For handling the effects, our implementation extends OCaml's |match ... with|
syntax with effect patterns. The expression
\begin{lstlisting}
match e with
| None -> false | Some b -> b
| effect (E s) k1 -> e1 | effect (F f) k2 -> e2
\end{lstlisting}
\noindent is translated to the equivalent of
\begin{lstlisting}
match_with (fun () -> e)
{ retc = (function None -> false | Some b -> b);
  effc = (function
  | (E s) -> (fun k1 -> e1)
  | (F f) -> (fun k2 -> e2)
  | e -> (fun k -> match perform e with
          | v -> continue k v
          | exception e -> discontinue k e)); }
\end{lstlisting}
For the sake of exposition, we introduce a |('a,'b) handler| type. This handler
handles a |'a comp| that returns a |'a| value, and itself returns a |'b| value.
The handler has a return case of type |'a -> 'b|. The effect case |effc|
handles effects of type |'c eff| with |('c,'b) continuation| and returns a
value of type |'b|. The last case in |effc| forwards any unmatched effect to
the outer handler and returns the value and exceptions back to the original
performer.

@@

Finally, we have the saved exception pointer, which points to the top-most
exception frame, and the saved stack pointer, which points to the top of the
stack. Switching between fiber only involves saving the exception and the stack
pointer of the current stack and loading the same on the target stack. Since
OCaml does not generate pointers into the stack, the two |fiber_info| fields
are the only ones that need to be updated when fibers are moved. We allocate
the space for the fiber using |malloc| and |free| it when no longer required.
We do not use a \emph{stack cache} of recently freed stacks for speeding up
fiber allocation. In our experiments, we observed that directly using
|mimalloc| outperforms having the stack cache.

@@

When we first enter OCaml from C, a new fiber is allocated for the main OCaml
stack. Since callbacks may be frequent in OCaml programs that use finalisers,
we run the callbacks on the same fiber as the current one. For example, the
layout of the Multicore OCaml stack at |caml_raise_exn| in the |meander|
example from \S\ref{sec:stack} is shown in Figure~\ref{fig:mcstack}. The
functions |caml_call_c| and |caml_call_ocaml| switch the stacks, and hence are
shown in both the system stack and the fiber. The details of the exception
frames have been elided in the figure. Thanks to the fiber representation,
external calls and callbacks remain fast.

\subsection{Effect handlers}

\begin{figure}
	\includegraphics[width=0.6\linewidth]{figures/fiber_handler}
	\caption{Layout of stack with effect handlers}
	\label{fig:fiber_handler}
\end{figure}

In order to illustrate the implementation of handling effects, consider the
example presented below:
\begin{lstlisting}
effect E : unit
effect F : unit;;
match (* comp_e *)
  match (* comp_f *) (*p1*) perform E (*p3*)
  with v -> v | effect F kf -> continue kf ()
with v -> v | effect E ke -> (*p2*) continue ke ()
\end{lstlisting}
The code performs |E| which is handled in the outer-most handler, and is
immediately resumed. The layout of the stack as the program executes is
captured in Figure~\ref{fig:fiber_handler}. At position |p1|, |rsp| is at the
top of the fiber |f|.

When an effect is performed, we invoke the effect closure |clos_heff| in the
fiber |f| with the current fiber pointer as the continuation. Since the first
handler does not handle effect |E|, the effect is forwarded to outer handler by
performing the effect closure of fiber |e|. This handles |E|. We allocate a one
word continuation object |ke| on the heap that holds the pointer to fiber |e|
which points to the fiber |f| (position |p2|). Thus, continuations are captured
without copying frames. Since each handler closure is evaluated until a machine
one is found, the take to handle an effect is linear in the number of handlers.

When the continuation is resumed, we overwrite the value of |ke| to |NULL| to
enforce at-most once semantics. Resuming a continuation involves traversing the
linked-list of fibers and making the last fiber point to the current fiber.
Just as in the operational semantics, the implementation invokes the
appropriate closure to either |continue| or |discontinue| the continuation
(position |p3|).
